\documentclass[12pt]{book}

%\usepackage{tgtermes}            % Times-like font from TeX Gyre
%\usepackage{tgpagella}           % Palatino-like font
%\usepackage{tgbonum}             % Bookman-like font
%\usepackage{bookman}
%\usepackage{libertine}           % Linux Libertine + math
%\usepackage{erewhon}             % Utopia alternative
\usepackage{fourier}             % Utopia text + Fourier math

\input{tex/packages.tex}

\begin{document}

\chapter{\lCalc}

In the early 20th century, mathematicians such as Bertrand Russell, David Hilbert, and Kurt Gödel were trying to set complete, consistent, and decidable foundations for mathematics. Within this context, Alonzo Church, in his effort to formalize the notion of  computability, introduced a minimal symbolic language based upon function abstraction and function application, this system is now called \lcalc.

In 1936, Church used lambda calculus to address Hilbert’s Entscheidungsproblem --whether a mechanical method could determine the truth of any first-order logic statement. He proved no such algorithm exists, establishing the undecidability of first-order logic.

One of the nuances formalized by the \lcalc \ is the distinction between extensional and intensional equality. The extensional approach to equivalence states that two functions are equivalent if they share input output pairs for every possible input. Its intensional counterpart extends this notion of equality by requiring that the procedures that compute these pairs share complexity i.e. they take the same steps toward yielding a result.\footnote{
  Let \( p \) be a sufficiently large prime, and let \( f, g : \mathbb{Z}_p \to \mathbb{Z}_p \) be defined by \( f(x) = x^2 \) and \( g(x) = \log_a(a^{x+2}) \), where \( a \in \mathbb{Z}_p^\times \) is a fixed primitive root. Although \( f \) and \( g \) are extensionally equal, i.e., they yield the same output for all inputs in \( \mathbb{Z}_p \), they are intensionally distinct. The function \( f \) performs a simple squaring operation, while \( g \) requires evaluating a discrete logarithm, intractable in general. \magenta{[Pedro Bonilla]}
}

Languages like LISP, Haskell, Erlang and others share the \lcalc \ as their theoretical foundation. The lambda calculus has established itself as the backbone of functional programming. The introduction of type systems into the \lcalc \ has allowed us to computer-verify mathematical proofs and develop programs that are correct by construction.\footnote{Principles of the \lcalc \ and Type Theory underlie every computer assisted verification tool as well as proof assistants and kitchen table programming languages like C or Java.}

\newpage
\input{tex/untyped-lambda-calc.tex}
\newpage
\input{tex/lambda-constructions.tex}
\newpage
\input{tex/fixed-point-recursion.tex}
\newpage
\input{tex/simply-typed-lambda-calc.tex}

\chapter{Intuitionistic Logic}
\newpage
\input{tex/intuitionistic-logic.tex}

\chapter{Category Theory}

Category theory abstracts the notion of internal structure and studies mathematical objects indirectly through their connections to other objects. In practical terms, instead of studying objects in isolation, category theory shifts the focus onto morphisms among objects, and that instead of studying mathematical entities from a definitional stand point, it emphasizes the importance of how constructions relate to one another.

The rulebook of categorical thinking keeps us away from manipulating mathematical objets directly, it also offers a categorical lens as a tool to look at constructions in a holistic spirit. This lens abstracts away the details of mathematical constructions but enables a more general approach to  problems. This perspective unifies many areas of mathematics under a common language.

Category theory offers extreme generality and has great power when it comes to unifying seemingly unrelated fields. Which often misleads people into interpreting it as a unifying theory of mathematics that has no practical applications, when in reality it is more like a tool used to create blueprints that describe mathematical environments and their properties. It is also commonly used to give structure to patterns that repeat within different fields of mathematics. In some instances, category theory serves as a medium for transporting mathematical statements—such as theorems or lemmas—between domains. A categorical formulation of such results can often be translated to other contexts via functors or equivalences of categories.

One of the fields that benefits from the practical applications of categorical reasoning is computer science, deeply related to category theory through type theory, intuitionistic logic and natural deduction. We only mention CCC's and LCCC's as relevant constructions in the effort not to give away the purpose of the chapter.

The general and unifying nature of categories brings us closer to the central focus of this work. In addition, most of what has been covered this far has a categorical representation that enables us to provide a unified approach to the already familiar type theory and intuitionistic logic. More on this in the following chapters.


\newpage
\input{tex/category-theory.tex}

\end{document}

